---
title: "PCA and Factor Analysis"
author: "cvieira"
date: "August 4, 2020"
output: html_document
---
  
  ```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Learning Outcomes
By the end of this Activity, you will be able to:
  
  * Invoke the kmeans algorithm to cluster individual students into groups

* Create different plots using the ggplot2 package to represent data from educational phenomena

*** 
  Part of this tutorial was adapted from:
  
  * http://r-statistics.co/Complete-Ggplot2-Tutorial-Part1-With-R-Code.html 

* Our own work on visualizations
+ Vieira, C., Parsons, P., & Byrd, V. (2018). Visual learning analytics of educational data: A systematic literature review and research agenda. Computers & Education, 122, 119-135.
+ Vieira, C., Magana, A. J., Roy, A., & Falk, M. L. (2019). Student explanations in the context of computational science and engineering education. Cognition and Instruction, 37(2), 201-231.
+ Magana, A. J., Fennell, H. W., Vieira, C., & Falk, M. L. (2019). Characterizing the interplay of cognitive and metacognitive knowledge in computational modeling and simulation practices. Journal of Engineering Education, 108(2), 276-303.
***
  
  Since we want to use the plots from the library ggplot2, the first thing we need to do is to import this library. 

```{r}
#install.packages('ggplot2')   # install it first if it is not installed yet
library(ggplot2)
library(tibble)

```

We also import the package tibble, which helps us with the kmeans clustering method.

#### Loading the Data


We first want to load some data, so we need to tell the computer where to get it from

Set the working directory to the folder where you have your files and read the file into a data frame

```{r}
#setwd('C:/Users/cvieira/Dropbox/Phd/Administrative/Mimi/FIE/2019/ScriptsAndData/data/')
setwd('C:/Users/cviei/Dropbox/Phd/Administrative/Mimi/FIE/2019/ScriptsAndData/data/')

typesOfKnowledge <- read.csv2("typesOfKnowledge.csv",header=TRUE,sep=",", row.names = 1)
meltedToK <- read.csv("meltedData.csv",header=TRUE,sep=",", row.names = 1)

```

### Activity 
How are these two data frames different? Check them out
```{r}
head(typesOfKnowledge)
head(meltedToK)

```

####Let's get to the clusters now...

We need to set a seed to always get the same set of clusters for the same data

And then we use kmeans function to identify four clusters in the data set.

```{r}
set.seed(1234)
stdClusters<- kmeans(typesOfKnowledge, 4)
stdClusters #Check out the cluster for each student
```


We need to transform the matrix into a data frame to map the clusters.

The as.factor function transforms a numerical variable into a categorical variable

```{r}
groups<- as.data.frame(typesOfKnowledge)

groups$stdClusters<- as.factor(stdClusters$cluster)
groups <- rownames_to_column(groups, "std") #The students are as row names and we need to have them as column to merge them later
```


Assign the clusters to each student in the melted format (which we will use to plot)

Make sure R recognizes the cluster as a categorical variable (it should not be a number!)
```{r}
meltedToK <- merge(meltedToK,groups[,c('std','stdClusters')], by.x = 'Student', by.y = 'std')
meltedToK$stdClusters <- factor(meltedToK$stdClusters, levels=c("1", "2","3","4"))
```

Now we can create the plot using ggplot.

Note that our X axis will have an interaction between the Types of Knowledge and the Steps of de modeling process (variable)
```{r}
ggplot(meltedToK, aes(x=interaction(ToK,variable),Student))+
  geom_point(aes(size=sum, shape=stdClusters))
```

The plot is ready, but it is better to see it organized, so we can do the following:
  
  1.  Get the list of students in order based on the clusters.
2.  Use that to organize the students in the data frame
```{r}
students <- groups[order(groups$stdClusters),]$std

meltedToK$Student <- factor(meltedToK$Student, levels=students)

#Plot it again:
ggplot(meltedToK, aes(x=interaction(ToK,variable),Student))+
  geom_point(aes(size=sum, shape=stdClusters))
```
### Activity 
How differet does our data look like if we choose three or four clusters?
  
  Hint:modify the call to kmeans, but make sure you start the script from the beginning because our structures have been modified.)

What else would you need to change? 
  
  
  
  
  ```{r}

```{r}




```{r}

```{r}




```{r}

```{r}


