---
title: "Tutorial Estad&iacute;stica B&aacute;sica y Modelo APG"
author: "cvieira"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE} 
library(learnr)
library(Hmisc)
#knitr::opts_chunk$set(echo = FALSE)
litReview <- read.csv2("../../datos/litReviewData.csv", header=TRUE, sep=";")
```

## Gr&aacute;ficas en GGPLot

### GGPlot

There are different ways to create plots in R. For example, you can create a histogram with the function 
*hist* . However, most of the plots created in R today are made using a library called *ggplot* . 
We need to first load *ggplot*

```{r loadLibrary, exercise=FALSE, exercise.eval=TRUE}
#install.packages('ggplot2')   # install it first if it is not installed yet

library(ggplot2)
```


We now want to load some data and our data is stored in a server. 

```{r readLItData, exercise=TRUE}

# The data is stored in a variable called litReview

litReview <- read.csv2("https://echidna.informaticaeducativa.cf/datos/litReviewData.csv", header=TRUE, sep=";")
```


Let's see the first rows our data set.

A quick reminder of our columns: 

* Connection to Visualization Background (CVG)

* Connection to Educational Theories (CET)

* Sophistication of the Visualization (SoV)

* Venue - Journal or conference where the paper was published

* Author - First author of the paper



#### Plotting

 The function ggplot will only create the blank plot specifying our axes

 For example, in the following instruction, we set up the CET as the x axis, and the SoV as the y-axis. 

 Check out what happens when you run it.

```{r createGGplot, exercise=TRUE}
ggplot(litReview, aes(x=CET, y=SoV))
```

###### Why did we include that function 'aes()' in the code?

```{r showHelp, exercise=TRUE}
?aes() # Describes how variables in the data are mapped to visual properties of geoms.
       # Everything we want to include into the visualization should be included there.
```

We now need to tell ggplot what kind of plot we want. 

```{r addGeomPoint, exercise=TRUE}
ggplot(litReview, aes(x=CET, y=SoV))+
  geom_point()  # Adding this line we are saying we want a scatter plot

```
###### Here are some other ideas:

* geom_point() ==> Scatter Plot

* geom_bar() ==> Bar Plot

* geom_line() ==> Line Plot

* geom_histogram() ==> Histogram

Each plot would require specific data and columns within the aes function. For instance, to plot a histogram, we only need x variable:

```{r addGHistogram, exercise=TRUE}
ggplot(litReview, aes(x=CET))+
  geom_histogram()
```

Here is a useful cheatsheet for different plots:

https://www.rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf

#### Activity 
 Change the plot to a line plot
```{r activity1, exercise=TRUE}
# write your code here.

```

Let's now get back to our scatter plot, which seems to be best for our purpose: 
*identifying gaps in this literature*

```{r backToPlotting, exercise=TRUE}
ggplot(litReview, aes(x=CET, y=SoV))+
      geom_point()  # Adding this line we are saying we want a scatter plot
```

These are still many points, and they are coming from differnt venues.
So, let's use the state to differentiate them with colors

```{r coloredPoints, exercise=TRUE}
ggplot(litReview, aes(x=CET, y=SoV))+
  geom_point(aes(color=Venue), size=3)  # Do you remember the aes() function? Here it is again
```

##### Aesthetics

You can change the color palette, edit the legend on the right, and many other things, but we will leave these aesthetics functions for some other time.
If you want to explore, here are a few functions you could use:

* theme(legend.position="None")  # remove legend

* scale_colour_brewer(palette = "Set1")  # change color palette


This is a nice simple scatter plot, but the scales in each axis seem odd. Let's adjust the axes to go on the whole range of our scale: 0-10
```{r settingAxis, exercise=TRUE}
ggplot(litReview, aes(x=CET, y=SoV))+
  geom_point(aes(color=Venue), size=3)+
  ylim(c(0, 10))
```

###### Did you notice that we just added up these new lines of code to the plot? 

That's how ggplot works, we can just continue adding new lines to modify our plot

##### Activity 

Set the limits 0 to 10 to the x-axis. 

```{r activity2, exercise=TRUE}
#Write your code here.

```

If we don't want to write the whole thing again, we can just store it in a variable
```{r saveToVariable, exercise=TRUE}

myPlot <- ggplot(litReview, aes(x=CET, y=SoV))+
  geom_point(aes(color=Venue), size=3)+
  ylim(c(0, 10))

```
  
And then, use that variable to add new things to our plot

```{r addTitles, exercise=TRUE}

myPlot+
  ggtitle("Two Dimensional Comparison", subtitle="Gap between Education and Visualization Researchers") + # Add a title and subtitle
  xlab("Connection to Edu. Theories") # Set the axis names

```
#### Activity 
 Set the name for the y-axis

```{r activity3, exercise=TRUE}
#Write your code here.

```

 We can also change the breaks in each axis. Here are a couple of parameters to explain


```{r addBreaksAxis, exercise=TRUE}

myPlot+
scale_y_continuous(breaks=seq(0, 10, 1), # Set a break in the y axis every unit
                   limits=c(0, 10)) # Set the limits
```

 Also note that we used scale_y_continuous because y is a continuous variable, but there are other types:

* scale_x_continuous, scale_y_continuous  ==> For numbers

* scale_x_discrete, scale_y_discrete   ==> For categories

* scale_x_date, scale_y_date   ==> For dates


#### Activity
 Change the x-axis limit so have breaks every 1

```{r activity4, exercise=TRUE}
#Write your code here.

```


 Now, there is an issue with our plot, there may be more than one point overlapping with each other at a single position.
 
 We can use the size as an indicaton of how many papers are at each point. To do that, we use geom_count instead of geom_point
 

```{r completePlot, exercise=TRUE}
myPlot <- ggplot(litReview, aes(x=CET, y=SoV))+
  geom_count(aes(color=Venue))+
  scale_x_continuous(breaks=seq(0, 10, 1), limits=c(0, 10))+
  scale_y_continuous(breaks=seq(0, 10, 1), limits=c(0, 10))+ 
  scale_size_continuous(breaks = seq(0, 5, 1))+
  labs(title="Two Dimensional Comparison", 
       subtitle="Gap between Education and Visualization Researchers",
       y="Sophistication of the Visualization", 
       x="Connection to Educational Theories", 
       size="# of Studies")
myPlot

```

#### Activity 


Create a scatter plot for *Connection to Educational Theories (CET) vs. Connectino to Visualizaiton Background (CET vs CVG)*

```{r activity5, exercise=TRUE}
#Write your code here.

```


